\section{Basic Assumptions}
Suppose a particle can be described by a probability distribution over space $x$ and time $t$, then the ``uniform'' particle is described by a plane wave
\begin{align}
\psi \propto \exp\left(
i~\dfrac{px-Et}{\hbar}
\right),
\end{align}
with constant momentum $p$ and energy $E$. The imaginary $i$ is needed to keep the amplitude of this wave $\psi^*\psi$ constant, whereas $\hbar$ is needed to remove units from the exponent. From derivatives of this plane wave, it is natural to conjecture the relations of conjugate variables to space and time, i.e. momentum and energy
\begin{align}
\left\{
\begin{array}{l}
\frac{d\psi}{dx} =  i\frac{p}{\hbar}\psi \\ [8 pt]
\frac{d\psi}{dt} = -i\frac{E}{\hbar}\psi
\end{array}
\right. \Rightarrow
\left\{
\begin{array}{l}
p\psi = -i\hbar\frac{d}{dx} \psi \\ [8 pt]
E\psi =  i\hbar\frac{d}{dt} \psi
\end{array}
\right..
\end{align}
Suppose total energy is a sum of non-relativist kinetic energy and local potential energy
\begin{align}
E = \frac{p^2}{2m} + V(x),
\end{align}
then the Hamiltonian operator
\begin{align}
\hat{\mathcal{H}} = \frac{\hat{p^2}}{2m} + V(x) = -\frac{\hbar^2}{2m}\nabla^2 + V(x).
\end{align}

\section{Helium and Jellium}
The quantum liquids helium and jellium are at the forefront of the quantum era ushered in the last century.
Having stable closed-shells, helium atoms interact mostly via an isotropic pair potential. It consists of a $r^{-6}$ van der Waals attraction at long range due to correlated dipole fluctuation and a hard-core repulsion upon electron density overlap~\cite{Aziz1979}.
Thus, the electronic problem is removed and only the ionic problem remains
\begin{align}
\hat{H}_{He} = \sum\limits_{I=1}^{N_I} -\dfrac{\hbar^2}{2m_I}\nabla_I^2
+\frac{1}{2}\sum\limits_{I=1}^{N_I}\sum\limits_{J=1,J\neq I}^{N_I} v(\vert\ri-\rj\vert).
\end{align}
In the opposite extreme, the jellium model replaces the material-dependent ionic potential in the electronic hamiltonian with a rigid homogeneous background of positive charge. In Hartree atomic units
\begin{align} \label{eq:intro-jellium}
\ham_{HEG} = \sum\limits_{i=1}^N -\dfrac{1}{2}\nabla_i^2 + \frac{1}{2}\sum\limits_{i=1}^N\sum\limits_{j=1,j\neq i}^N
\dfrac{1}{\vert \bs{r}_i-\bs{r}_j \vert}
+E_{e-b}+E_{b-b},
\end{align}
where $E_{e-b}$ and $E_{b-b}$ are constants due to electron-background and background-background interactions.

The jellium model eq.~(\ref{eq:intro-jellium}) brings the electronic problem into focus. The isotropic background eliminates potential symmetry breaking interactions that can be introduced by a crystalline arrangement of the ions. The rigidity of the background also removes electron-ion coupling effects. This model was studied in great detail in the past century and its ground-state behavior was largely understood. Much progress has even been made regarding its excitations and finite-temperature properties.

There is only one length scale in the jellium model: the average electron-electron separation $r_s=a/a_B$
\begin{align}
\dfrac{1}{n} \equiv \dfrac{\Omega}{N} = \left\{\begin{array}{lr}
2 r_s a_B & 1D\\
\pi(r_sa_B)^2 & 2D\\
\frac{4\pi}{3}(r_sa_B)^3 & 3D
\end{array}\right.,
\end{align}
where $a_B$ is the Bohr radius. The kinetic energy scales as $r_s^{-2}$ (due to $\nabla^2$) while the potential energies scales as $r_s^{-1}$ (for $1/r$ potential), so $r_s$ measures the relative strength of potential to kinetic energy. In this sense, $r_s$ is the zero-temperature analogue of the classical coulomb coupling parameter
\begin{align}
\Gamma \equiv \dfrac{e^2n^{1/d}}{k_BT}.
\end{align}
The valence electron density in alkaline metals is $r_s\sim2$, meaning the kinetic energy is important, so the electrons delocalize and form a liquid to minimize kinetic energy. At sufficiently large $r_s$ ($\sim 100$ in 3D and $\sim 30$ in 2D), the potential energy dominates, so the electrons localize to form a Wigner crystal.

%\subsection{Exact Electron-Ion Schr\"odinger equations towards Born-Oppenheimer}


\begin{comment}
\begin{table}[h]
\centering
\begin{tabular}{cccccc}
\toprule
     & Temperature & Classical & Quantum & Quantization & Sign Problem \\
\midrule
PIMC &     any     &   yes     &   yes   & first & $\braket{\sigma}\propto \exp\left[ -\beta N (F_f-F_b) \right]$ \\
DMC  &    zero     &    no     &   yes   & first & $\braket{\sigma}\propto \exp\left[ -\beta N (F_f-F_T) \right]$ \\
FP-DMC & zero & no & yes & first & no \\
VMC & low & no & yes & first & no \\
AFQMC & low & no & yes & second & ? \\
\bottomrule
\end{tabular}
\end{table}
\end{comment}