\listoftables
\listoffigures

%% Create a List of Abbreviations. The left column
%% is 1 inch wide and left-justified
\chapter{List of Abbreviations}

\begin{symbollist*}
\item[BOA] Born-Oppenheimer Approximation.
\item[DBOC] Diagonal Born-Oppenheimer Correction.
\item[EOS] Equation Of State.
\item[DM] Density Matrix.
\item[QMC] Quantum Monte Carlo.
\item[RPIMC] Restricted Path Integral Monte Carlo.
\item[VPI] Variational Path Integral.
\item[RMC] Reptation Monte Carlo
\item[FP-DMC] Fixed-Phase Diffusion Monte Carlo.
\item[VMC] Variational Monte Carlo.
\item[KW] Kolos and Wolniewicz.
\item[RHF] Restricted Hartree Fock.
\item[CISD] Configuration Interaction Singles Doubles.
\item[CAS] Complete Active Space.
\item[ANO] Atomic Natural Orbital.
\item[ECG] Explicitly-Correlated Gaussian.
\item[KS-DFT] Kohn-Sham Density Functional Theory.
\item[HK] Hohenberg and Kohn.
\item[LDA] Local Density Approximation.
\item[PZ] Perdew and Zunger.
\item[HOMO] Highest-Occupied Molecular Orbital.
\item[LUMO] Lowest-Unoccupied Molecular Orbital.
\item[PBE] Perdew, Burke, and Ernzerhof.
\item[HSE] Heyd, Scuseria, and Ernzerhof.
\item[SBJ] Slater-Backflow-Jastrow.
\item[RPA] Random Phase Approximation.
\item[RDM] Reduced Density Matrix.
\item[LLT] Liquid-Liquid Transition.
\item[DAC] Diamond Anvil Cell.
\item[MD] Molecular Dynamics.
\item[PES] Potential Energy Surface.
\item[GPa] Giga-Pascal
\item[LP] Low-pressure Phase.
\item[BSP] Broken-Symmetry Phase.
\item[IR] InfraRed.
\item[EQQ] Electric Quadrupole-Quadrupole.
\item[GCTABC] Grand-Canonical Twist Average Boundary Condition.
\item[KZK] Kwee, Zhang, and Krakauer.
\item[MSD] Mean-Squared Deviation.
\item[FWHM] Full Width at Half Max.
\item[SDF] Spectral Density Function.
\item[BFD] Burkatzki, Filippi, and Dolg.
\item[QE] Quantum Espresso.
\item[PPC] PseudoPotential Correction.
\item[FSC] Finite-Size Correction.
\item[BF] Back Flow.
\item[ZPE] Zero-Point Energy.
\end{symbollist*}

%% Create a List of Symbols. The left column
%% is 0.7 inch wide and centered
\chapter{List of Symbols}

\begin{symbollist}[0.7in]
\item[$h$] Planck constant $\hbar=\dfrac{h}{2\pi}$.
\item[$\Psi$] many-body electron-ion wave function.
\item[$\psi$] many-body electron wave function.
\item[$\phi$] one-body electron wave function.
\item[$\chi$] one-body ion wave function.
\item[$\bs{r}_i$] 3D position of electron $i$.
\item[$\bs{r}_I$] 3D position of ion $I$.
\item[$\bs{R}$] positions of all $N$ electrons.
\item[$\bs{R}_I$] positions of all $N_I$ ions.
\item[$\rho$] density matrix or electron density.
\item[$N$] number of electrons.
\item[$N_I$] number of ions.
\item[$\Omega$] total volume.
\item[$L$] linear extent of the simulation cell.
\item[$r_s$] Wigner-Seitz radius.
\item[$k_F$] Fermi wave vector.
\item[$\Gamma$] classical coulomb coupling parameter.
\item[$\beta$] inverse temperature.
\item[$t$] imaginary time unless otherwise specified.
\item[$\tau$] imaginary time step unless otherwise specified.
\item[$Z$] canonical partition function.
\item[$\hat{H}$] electron-ion hamiltonian.
\item[$\ham$] electronic hamiltonian.
\item[$\lambda$] quantumness $\lambda=\dfrac{\hbar^2}{2m}$.
\item[$\Lambda_{kj}$] nonadiabatic coupling between electronic states $k$ and $j$.
\item[$\Lambda$] de Broglie wavelength.
\item[$P$] pressure.
\item[$V$] total potential energy.
\item[$T$] total kinetic energy.
\item[$E$] total energy.
\item[$E_T$] trial total energy.
\item[$\Psi_T$] trial wave function.
\item[$\sigma^2$] variance of the local energy.
\item[$\pi$] probability distribution.
\item[$\Delta$] band gap.
\item[$g(r)$] pair correlation function.
\item[$S(k)$] static structure factor.
\item[$U$] negative exponent of the Jastrow wave function.
\item[$u(r)$] pair contribution to $U$.
\item[$\omega_p$] plasmon frequency.
\item[$\epsilon_{\bs{k}}$] dielectric function.
\item[$n(\bs{k})$] momentum distribution.
\item[$J(p)$] Compton profile.
\end{symbollist}